---
title: "SNP_PFAS_riskanalysis"
author: "Aina Vaivade"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    fig_caption: yes
params:
  resultDir: '/Users/ainva234/Documents/Project/EIMS/Results/Environmental/Targeted/Genes*PFAS/'
  emisColestVisit: '/Users/ainva234/Documents/Project/EIMS/Data/EIMS_closest_visit.csv'
  eimsSMSReg: '/Users/ainva234/Documents/Project/EIMS/Data/SMSRegData/EIMS_Criteria_smsreg_incl_drugs.csv'
  dataDir: '/Users/ainva234/Documents/Project/EIMS/Data/'
  funcFile: '/Users/ainva234/Documents/Project/EIMS/Code/functions.R'
  samplingData: '/Users/ainva234/Documents/Project/EIMS/Data/EIMS_collection.xlsx'
  lifeStyle: "~/Documents/Project/EIMS/Data/survey_data/documents_20230619/EIMS_metabolomics_LifestyleExposures.txt"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      messeges = FALSE)
```

```{r Libraries}
# Load libraries
install.packages("readxl")
install.packages("lme4")
install.packages("car")
install.packages("dplyr")
install.packages("purrr")

library(readxl)
library(lme4)
library(car)
library(dplyr)
library(purrr)
```

```{r Load data}
source(params$funcFile)

load(paste0(params$dataDir, "Environmental/Targeted/PFAS/dt.pfas.20cov.RData"))
dt.closest.visit <- read.csv(params$emisColestVisit)
eims.metadata <- read.csv(paste0(params$dataDir, "collated_eims_metadata.csv"))
eims.smsreg.key <- read_excel(paste0(params$dataDir, "SMS_reg_inclEIMS.xlsx"))
smsReg <- read.csv(params$eimsSMSReg)
compound.info <- read_excel(paste0(params$dataDir, 'Environmental/ENV_TARGETED_COMPOUNDS_INFO.xlsx'))
runorder <- read_excel(paste0(params$dataDir, "Environmental/FinalRunOrderEIMS_Enviro2022.xlsx"), sheet = "Runorder")
dt.sampling <- read_excel(params$samplingData)
dt.lifeStyle <- read.csv(params$lifeStyle, sep = "\t")
dt.children <- read_excel(paste0(params$dataDir, "/survey_data/documents_20240419/children.xls"))
snp.data <- read_excel(paste0(params$dataDir, "SNP/Burman_MSrisk_230205.xlsx"))

dt.pfas <- dt.cov
```

# Order accordingly size (number of C)
```{r order data to same order}
colnames(dt.pfas)[which(colnames(dt.pfas) == "PFHps")] <- "PFHpS"
colnames(dt.pfas)[which(colnames(dt.pfas) == "PCB 187")] <- "4-OH-CB187"
colnames(dt.pfas)[which(colnames(dt.pfas) == "PCB 153")] <- "3-OH-CB153"
colnames(dt.pfas)[which(colnames(dt.pfas) == "4-Hydroxy-PCB 106")] <- "4-OH-CB107"

order <- c("PFBS", "PFPeS", "PFHpA",
           "PFHxS", "PFHpS", "PFOA",
           "PFOS", "FOSA", "N-MeFOSAA",
           "N-EtFOSAA", "PFNA", "PFDA",
           "PFDS", "PFUnA", "PFDoA",
           "PFTrDA", "8:2 FTS", "4-OH-CB187",
           "3-OH-CB153", "4-OH-CB107",
           "Type", "pairID", "Age",
           "Sex", "YOB", "County",
           "Week", "time.diff.sampling",
           "diag.samp.diff", "BMI_20", "BMI_curr")

dt.pfas <- dt.pfas[,order]
```

```{r Impute NA or negative values}
dt.pfas.imp <- dt.pfas
dt.pfas.imp[,1:20] <- 2^dt.pfas.imp[,1:20]

# Impute !/2 in vaue for NA and negative values
for(i in 1:20){
  if(length(which(dt.pfas.imp[,i] < 0)) > 0){
    dt.pfas.imp[which(dt.pfas.imp[,i] < 0),i] <- NA
  }
  low.val <- min(dt.pfas.imp[,i], na.rm=T)
  dt.pfas.imp[which(is.na(dt.pfas.imp[,i])),i] <- low.val/2
  dt.pfas.imp[,i] <- log(dt.pfas.imp[,i], base = 2)
}

dt.pfas <- dt.pfas.imp
```

# Include Lifestyle data
  - Reegular smoking
  - Passive smoking
  - Irregular smoking
  - Swedish snuff
  - Mono infection
  - Alcohol consumption
  - Number of childbirths
  - Sun-habits 
  - Residential area
  - Born in Sweden or other Nordic countries

```{r lifestyle data}
dt.pfas$collection.year <- NA

for(i in rownames(dt.pfas)){
  dt.pfas$collection.year[which(rownames(dt.pfas) == i)] <- format(as.Date(dt.sampling$collectiondate[which(dt.sampling$cdk == eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == i)])]), "%Y")
}

# Set an index date which
## The index date is the same as diagnosis date, and for HC the index date is the diagnosis date of their matched MS subject
dt.pfas$index.date <- NA
for(i in 1:nrow(dt.pfas)){
  if(dt.pfas$Type[i]!= "HC"){
    eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
    pat.code <- as.numeric(strsplit(eims.smsreg.key$Register.Id[which(eims.smsreg.key$EIMS_1 == eims.id)], "SMS")[[1]][2])
    temp <- smsReg %>% filter(patient.code == pat.code)
    diag.date <- temp$diagnosis_date
    if(!is.na(min(as.Date(diag.date)))){
      dt.pfas$index.date[i] <- as.character(min(as.Date(diag.date)))
    }else{ #If diagnosis date is NA, the debute date is used
      dt.pfas$index.date[i] <- as.character(min(as.Date(temp$debut_date)))
    }
  }
}

for(i in 1:nrow(dt.pfas)){
  if(dt.pfas$Type[i] == "HC"){
    dt.pfas$index.date[i] <- dt.pfas$index.date[which(dt.pfas$pairID == dt.pfas$pairID[i] & dt.pfas$Type != "HC")]
  }
}

# Include regular smoking
dt.lifestyle <- read.csv("/Users/ainva234/Documents/Project/EIMS/Data/survey_data/documents_20230619/EIMS_metabolomics_LifestyleExposures.csv", sep = ";")
dt.smoking <- read.csv("/Users/ainva234/Documents/Project/EIMS/Data/survey_data/documents_20230619/EIMS_metabolomics_regularSmoke.csv", sep = ";")

dt.pfas$reg.smoking <- NA

for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
  if(is.na(dt.lifestyle$neverSmoke[which(dt.lifestyle$EIMS == eims.id)])){
      dt.pfas$reg.smoking[i] <- NA
  }else if(dt.lifestyle$neverSmoke[which(dt.lifestyle$EIMS == eims.id)] == 2){ #Never smoked
      dt.pfas$reg.smoking[i] <- 0
  }else{ #Smoker or has been. Check if the individual was a smoker at or prior to the index year
      smoking <- dt.smoking[which(dt.smoking$EIMS == eims.id),]
      start.smoking.age <- min(smoking$fromAge)
      if(!is.na(start.smoking.age)){
        sample.coll.year <- eims.metadata$year[which(eims.metadata$EIMS_ID == eims.id)]
        start.smoking.year <- sample.coll.year - (dt.pfas$Age[i] - start.smoking.age)
        if(is.na(dt.pfas$index.date[i])){
          dt.pfas$reg.smoking[i] <- NA
        }else if(as.numeric(start.smoking.year) <= as.numeric(format(as.Date(dt.pfas$index.date[i]), "%Y"))){
          end.smoking.age <- max(smoking$toAge)
          if(is.na(end.smoking.age)){
            dt.pfas$reg.smoking[i] <- 1
          }else if(end.smoking.age >= dt.pfas$Age[i]){
            dt.pfas$reg.smoking[i] <- 1
          }else{
            dt.pfas$reg.smoking[i] <- 0
          }
        }else{
          dt.pfas$reg.smoking[i] <- 0
        }
      }else{
        dt.pfas$reg.smoking[i] <- NA
      }
  }
}

dt.pfas$reg.smoking <- relevel(as.factor(dt.pfas$reg.smoking), ref = "0")
rm(dt.smoking)

# Include irregular smoking
dt.smoking <- read.csv("/Users/ainva234/Documents/Project/EIMS/Data/survey_data/documents_20230619/EIMS_metabolomics_irregSmoke.csv", sep = ";")

dt.pfas$irreg.smoking <- NA

for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
  if(is.na(dt.lifestyle$irregSmoke[which(dt.lifestyle$EIMS == eims.id)])){
      dt.pfas$irreg.smoking[i] <- NA
  }else if(dt.lifestyle$irregSmoke[which(dt.lifestyle$EIMS == eims.id)] == 1){ #Never smoked
      dt.pfas$irreg.smoking[i] <- 0
  }else{ #Smoker or has been. Check if the individual was a smoker at or prior to the index year
      smoking <- dt.smoking[which(dt.smoking$EIMS == eims.id),]
      start.smoking.age <- min(smoking$fromAge)
      if(!is.na(start.smoking.age)){
        sample.coll.year <- eims.metadata$year[which(eims.metadata$EIMS_ID == eims.id)]
        start.smoking.year <- sample.coll.year - (dt.pfas$Age[i] - start.smoking.age)
        if(is.na(dt.pfas$index.date[i])){
          dt.pfas$irreg.smoking[i] <- NA
        }else if(as.numeric(start.smoking.year) <= as.numeric(format(as.Date(dt.pfas$index.date[i]), "%Y"))){
          end.smoking.age <- max(smoking$toAge)
          if(is.na(end.smoking.age)){
            dt.pfas$irreg.smoking[i] <- 1
          }else if(end.smoking.age >= dt.pfas$Age[i]){
            dt.pfas$irreg.smoking[i] <- 1
          }else{
            dt.pfas$irreg.smoking[i] <- 0
          }
        }else{
          dt.pfas$irreg.smoking[i] <- 0
        }
      }else{
        dt.pfas$irreg.smoking[i] <- NA
      }
  }
}

dt.pfas$irreg.smoking <- relevel(as.factor(dt.pfas$irreg.smoking), ref = "0")
rm(dt.smoking)

# Include passive smoking (sum work and home)
dt.passive.home <- read.csv("/Users/ainva234/Documents/Project/EIMS/Data/survey_data/documents_20230619/EIMS_metabolomics_passSmokeHome.csv", sep = ";")
dt.passive.work <- read.csv("/Users/ainva234/Documents/Project/EIMS/Data/survey_data/documents_20230619/EIMS_metabolomics_passSmokeWork.csv", sep = ";")

dt.pfas$passive.smoking.home <- NA
dt.pfas$passive.smoking.work <- NA

for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas)[i])]
    
  # passive smoking in the home environment 
  fromYear.home <- NA
  toYear.home <- NA  
  if(is.na(dt.lifestyle$passSmokeHome[which(dt.lifestyle$EIMS == eims.id)])){
      dt.pfas$passive.smoking.home[i] <- NA
  }else if(dt.lifestyle$passSmokeHome[which(dt.lifestyle$EIMS == eims.id)] == 1){
     dt.pfas$passive.smoking.home[i] <- 0 
  }else{
      smoking <- dt.passive.home[which(dt.passive.home$EIMS == eims.id),]
      if(!is.na(min(smoking$fromYear))){
        sample.coll.year <- eims.metadata$year[which(eims.metadata$EIMS_ID == eims.id)]
        if(is.na(dt.pfas$index.date[i])){
          dt.pfas$passive.smoking.home[i] <- NA
        }else if(as.numeric(min(smoking$fromYear)) <= as.numeric(format(as.Date(dt.pfas$index.date[i]), "%Y"))){
          if(is.na(max(smoking$toYear))){
            dt.pfas$passive.smoking.home[i] <- 1
          }else if(max(smoking$toYear) >= as.numeric(format(as.Date(dt.pfas$index.date[i]), "%Y"))){
            dt.pfas$passive.smoking.home[i] <- 1
          }else{
            dt.pfas$passive.smoking.home[i] <- 0
          }
        }else{
          dt.pfas$passive.smoking.home[i] <- 0
        }
      }else{
        dt.pfas$passive.smoking.home[i] <- NA
      }
  }

  # Passive smoking in the working environment
  fromYear.work <- NA
  toYear.work <- NA    
  if(is.na(dt.lifestyle$passSmokeWork[which(dt.lifestyle$EIMS == eims.id)])){
      dt.pfas$passive.smoking.work[i] <- NA
  }else if(dt.lifestyle$passSmokeWork[which(dt.lifestyle$EIMS == eims.id)] == 1){
     dt.pfas$passive.smoking.work[i] <- 0 
  }else{
      smoking <- dt.passive.work[which(dt.passive.work$EIMS == eims.id),]
      if(!is.na(min(smoking$fromYear))){
        sample.coll.year <- eims.metadata$year[which(eims.metadata$EIMS_ID == eims.id)]
        if(is.na(dt.pfas$index.date[i])){
          dt.pfas$passive.smoking.work[i] <- NA
        }else if(as.numeric(min(smoking$fromYear)) <= as.numeric(format(as.Date(dt.pfas$index.date[i]), "%Y"))){
          if(is.na(max(smoking$toYear))){
            dt.pfas$passive.smoking.work[i] <- 1
          }else if(max(smoking$toYear) >= as.numeric(format(as.Date(dt.pfas$index.date[i]), "%Y"))){
            dt.pfas$passive.smoking.work[i] <- 1
          }else{
            dt.pfas$passive.smoking.work[i] <- 0
          }
        }else{
          dt.pfas$passive.smoking.work[i] <- 0
        }
      }else{
        dt.pfas$passive.smoking.work[i] <- NA
      }
  }
}

dt.pfas$passive.smoking <- NA
for(i in 1:nrow(dt.pfas)){
  if(is.na(dt.pfas$passive.smoking.work[i])){
    dt.pfas$passive.smoking[i] <- dt.pfas$passive.smoking.home[i]
  }else if(!is.na(dt.pfas$passive.smoking.work[i]) & !is.na(dt.pfas$passive.smoking.home[i])){
    if(dt.pfas$passive.smoking.home[i] == 0 & dt.pfas$passive.smoking.work[i] != 1){
      dt.pfas$passive.smoking[i] <- 0
    }else if(dt.pfas$passive.smoking.home[i] == 1){
      dt.pfas$passive.smoking[i] <- 1
    }
  }
  
  if(is.na(dt.pfas$passive.smoking.home[i])){
    dt.pfas$passive.smoking[i] <- dt.pfas$passive.smoking.work[i]
  } else if(!is.na(dt.pfas$passive.smoking.home[i]) & !is.na(dt.pfas$passive.smoking.work[i])){
      if(dt.pfas$passive.smoking.work[i] == 0 & dt.pfas$passive.smoking.home[i] != 1){
        dt.pfas$passive.smoking[i] <- 0
      }else if(dt.pfas$passive.smoking.work[i] == 1){
        dt.pfas$passive.smoking[i] <- 1
      }
    }
  
}
dt.pfas$passive.smoking <- relevel(as.factor(dt.pfas$passive.smoking), ref = "0")
rm(dt.passive.home, dt.passive.work)

# Include Swedish snuff
dt.snus <- read.csv("/Users/ainva234/Documents/Project/EIMS/Data/survey_data/documents_20230619/EIMS_metabolomics_Snus.csv", sep = ";")

dt.pfas$snuff <- NA

for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
  if(is.na(dt.lifestyle$Snus[which(dt.lifestyle$EIMS == eims.id)])){
      dt.pfas$snuff[i] <- NA
  }else if(dt.lifestyle$Snus[which(dt.lifestyle$EIMS == eims.id)] == 1){
      dt.pfas$snuff[i] <- 0
  }else{ 
      snus <- dt.snus[which(dt.snus$EIMS == eims.id),]  
      start.snus.age <- min(dt.snus$fromAge)
      if(!is.na(start.snus.age)){
        sample.coll.year <- eims.metadata$year[which(eims.metadata$EIMS_ID == eims.id)]
        start.snus.year <- sample.coll.year - (dt.pfas$Age[i] - start.snus.age)
        if(is.na(dt.pfas$index.date[i])){
          dt.pfas$snuff[i] <- NA
        }else if(as.numeric(start.snus.year) <= as.numeric(format(as.Date(dt.pfas$index.date[i]), "%Y"))){
          end.snus.age <- max(dt.snus$toAge)
          if(is.na(end.snus.age)){
            dt.pfas$snuff[i] <- 1
          }else if(end.snus.age >= dt.pfas$Age[i]){
            dt.pfas$snuff[i] <- 1
          }else{
            dt.pfas$snuff[i] <- 0
          }
        }else{
          dt.pfas$snuff[i] <- 0
        }
      }else{
        dt.pfas$snuff[i] <- NA
      }
  }
}                    

dt.pfas$snuff <- as.factor(dt.pfas$snuff)
rm(dt.snus)

# Include tratment
dt.drugs <- read.csv("/Users/ainva234/Documents/Project/EIMS/Data/SMSRegData/terapi.csv", sep=";")
drug.classification <- read_excel("/Users/ainva234/Documents/Project/EIMS/Data/SMSRegData/drug_classification_V2.xlsx")

# Inlcude mono
dt.pfas$mono <- NA

for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
  if(is.na(dt.lifestyle$infMono[which(dt.lifestyle$EIMS == eims.id)])){
      dt.pfas$mono[i] <- NA
  }else if(dt.lifestyle$infMono[which(dt.lifestyle$EIMS == eims.id)] == 1){
      dt.pfas$mono[i] <- 0
  }else{ 
      dt.pfas$mono[i] <- 1
  }
} 
dt.pfas$mono <- relevel(as.factor(dt.pfas$mono), ref = "0")

# Include cl alcohol consumed passed week and if the alcohol consumption has changed comapred to 5 years ago
dt.pfas$cl.alcohol <- NA
dt.pfas$alcohol5Years <- NA

dt.lifestyle$alcoholBeerQuant <- dt.lifestyle$alcoholBeerQuant*0.035/40
dt.lifestyle$alcoholWineQuant <- dt.lifestyle$alcoholWineQuant*0.12/40
dt.lifestyle$alcoholStrongWineQuant <- dt.lifestyle$alcoholStrongWineQuant*0.15/40
dt.lifestyle$alcoholLiquorQuant <- dt.lifestyle$alcoholLiquorQuant

for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
  if(is.na(dt.lifestyle$alcohol12Months[which(dt.lifestyle$EIMS == eims.id)])){
      dt.pfas$cl.alcohol[i] <- NA
  }else if(dt.lifestyle$alcohol12Months[which(dt.lifestyle$EIMS == eims.id)] == 1){
      dt.pfas$cl.alcohol[i] <- 0
  }else{ 
      dt.pfas$cl.alcohol[i] <- sum(dt.lifestyle[which(dt.lifestyle$EIMS == eims.id),c(30:33)], na.rm=T) 
  }
  dt.pfas$alcohol5Years[i] <- dt.lifestyle$alcohol5Year[which(dt.lifestyle$EIMS == eims.id)]
} 

dt.pfas$alcohol5Years <- as.factor(dt.pfas$alcohol5Years)

# Include number of childbirths prior to sample colleciton
dt.pfas$nr.children <- NA
for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
  dt.pfas$nr.children[i] <- dt.children$antalbarn[which(dt.children$EIMS == eims.id)]
} 
dt.pfas$nr.children[which(dt.pfas$Sex == 0)] <- 0
dt.pfas$nr.children[which(is.na(dt.pfas$nr.children))] <- 0

# Assumption regarding BMI
## Patients in the span of 15 to 25 years are assumed to ahve the same BMI at 20 as the current BMI and vice versa
dt.pfas$BMI_20[which(dt.pfas$Age %in% c(15:25) & is.na(dt.pfas$BMI_20))] <- dt.pfas$BMI_curr[which(dt.pfas$Age %in% c(15:25) & is.na(dt.pfas$BMI_20))]
dt.pfas$BMI_curr[which(dt.pfas$Age %in% c(15:25) & is.na(dt.pfas$BMI_curr))] <- dt.pfas$BMI_20[which(dt.pfas$Age %in% c(15:25) & is.na(dt.pfas$BMI_curr))]

# Include residential area
dt.pfas.MS <- dt.pfas[which(dt.pfas$Type != "HC"),]
for(i in unique(dt.pfas$pairID)){
  if(length(which(dt.pfas.MS$pairID == i)) > 0){
   dt.pfas$County[which(dt.pfas$pairID == i)] <- dt.pfas.MS$County[which(dt.pfas.MS$pairID == i)]
  }
}

# Sun habits
## Sum up the three questions regarding sun habits (min score = 3 and max score = 12)
## Group in sun habits (low < 25% qutlie, medium and high > 75% quatile) based on controls sunhabits
dt.pfas$sunhabit.score <- NA
for (i in 1:nrow(dt.pfas)) {
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas )[i])]
  dt.pfas$sunhabit.score[i] <- sum(c(dt.lifestyle$sunSolarium[which(dt.lifestyle$EIMS == eims.id)],
                                     dt.lifestyle$sunTravel[which(dt.lifestyle$EIMS == eims.id)],
                                     dt.lifestyle$sunTan[which(dt.lifestyle$EIMS == eims.id)]), na.rm=T)
}
low <- quantile(dt.pfas$sunhabit.score[which(dt.pfas$Type == "HC")], probs = c(0.25, 0.75))[1]
high <- quantile(dt.pfas$sunhabit.score[which(dt.pfas$Type == "HC")], probs = c(0.25, 0.75))[2]

dt.pfas$sunhabit[which(dt.pfas$sunhabit.score <= low)] <- "low"
dt.pfas$sunhabit[which(dt.pfas$sunhabit.score >= high)] <- "high"
dt.pfas$sunhabit[which(dt.pfas$sunhabit.score > low & dt.pfas$sunhabit.score < high)] <- "medium"

dt.pfas$sunhabit <- relevel(as.factor(dt.pfas$sunhabit), ref = "low")

# Born in Sweden and other nordic countries
born.country <- read_excel("/Users/ainva234/Documents/Project/EIMS/Data/survey_data/Ancestry.xls")
dt.pfas$Swedish <- NA
dt.pfas$Nordic <- NA
for(i in 1:nrow(dt.pfas)){
  eims.id <- eims.metadata$EIMS_ID[which(eims.metadata$eims_filename == rownames(dt.pfas)[i])]
  dt.pfas$Swedish[i] <- born.country$swedish[which(born.country$EIMS == eims.id)]
  dt.pfas$Nordic[i] <- born.country$nordic[which(born.country$EIMS == eims.id)]
}
dt.pfas$Swedish <- as.factor(dt.pfas$Swedish)
dt.pfas$Nordic <- as.factor(dt.pfas$Nordic)
```

# Logistic regression
Is PFAS exposure a risk factor for developing MS

```{r set status (1/0)}
#Set disease status, 1 for individuals with MS and 0 for controls
dt.pfas$status <- 1 
dt.pfas$status[which(dt.pfas$Type == "HC")] <- 0
dt.pfas$status <- relevel(as.factor(dt.pfas$status), ref = "0")
dt.pfas$Sex <- relevel(as.factor(dt.pfas$Sex), ref = "0")
```

## Base model
```{r Males and Females}
covariate.table <- matrix(NA, nrow = 20, ncol = 9)
rownames(covariate.table) <- colnames(dt.pfas)[1:20]
colnames(covariate.table) <- c("Age - estimate", "Age - P",
                               "Sex - estimate", "Sex - P",
                               "Current BMI - estimate", "Current BMI - P",
                               "BMI at 20 - estimate", "BMI at 20 - P",
                               "Week - P")

summary.PFAS <- matrix(NA, nrow = 20, ncol = 3)
colnames(summary.PFAS) <- c("OR", "CI 95%", "P-value")
rownames(summary.PFAS) <- colnames(dt.pfas)[1:20]

for(i in 1:20){
  temp <- dt.pfas %>% dplyr::select(append(colnames(dt.pfas)[i], c("Age", "Sex", "BMI_curr", "BMI_20", "Week", "status", "County")))
  names(temp)[1] <- "envCont"
  mdl <- glmer(status ~ envCont + Age + Sex + BMI_curr + BMI_20 + Week + (1|County), data = temp, family = binomial(link='logit'))
  
  # Odds ratios, 95% CI, and p-values for all covariates in the model
  conf.int <- Confint(mdl, exponentiate = TRUE)
  summary.mdl <- S(mdl, exponentiate = T)
  mdl.summary <- cbind(round(exp(summary(mdl)$coefficients[2:7,1]), digits = 3),
                       cbind(paste(round(exp(conf.int)[3:8,1], digits = 3), "-", round(exp(conf.int)[3:8,2], digits = 3)),  
                             round(summary.mdl$coefficients[2:7,4], digits =3)))
  colnames(mdl.summary) <- c("OR", "CI 95%", "P-value")
  
  # PFAS/OH-PCBs OR, 95%CI, and p-value
  summary.PFAS[i,] <- c(round(exp(summary(mdl)$coefficients[2]), digits = 3),
                       paste(round(exp(conf.int)[3,1], digits = 3), "-", round(exp(conf.int)[3,2], digits = 3)), 
                       round(coef(summary(mdl))[2,4], digits =3))
  
  # Covariates estimates and Anova p-values
  covariate.table[i,c(1,3,5,7)] <- format(signif(summary(mdl)$coefficients[3:6], digits = 3), scientific = T)
  covariate.table[i,c(2,4,6,8:9)] <- format(signif(Anova(mdl)[2:6,3], digits = 3), scientific = T)
}

write.csv(summary.PFAS, file = paste0(params$resultDir, "LogisticRegression/BaseModel/summary.compounds.csv"))
write.csv(covariate.table, file = paste0(params$resultDir, "LogisticRegression/BaseModel/covariates.csv"))
```

### Divided by Sex
```{r  Males, message=FALSE}
covariate.table <- matrix(NA, nrow = 20, ncol = 7)
rownames(covariate.table) <- colnames(dt.pfas)[1:20]
colnames(covariate.table) <- c("Age - estimate", "Age - P",
                               "Current BMI - estimate", "Current BMI - P",
                               "BMI at 20 - estimate", "BMI at 20 - P",
                               "Week - P")

summary.PFAS <- matrix(NA, nrow = 20, ncol = 3)
colnames(summary.PFAS) <- c("OR", "CI 95%", "P-value")
rownames(summary.PFAS) <- colnames(dt.pfas)[1:20]

for(i in 1:20){
  temp <- dt.pfas %>% dplyr::select(append(colnames(dt.pfas)[i], c("Age", "Sex", "BMI_curr", "BMI_20", "Week", "status")))
  
  temp <- temp[which(temp$Sex == 0),] # Select only males
  names(temp)[1] <- "envCont"
  mdl <- glm(status ~ envCont + Age + BMI_curr + BMI_20 + Week, 
             data = temp, 
             family = binomial)
  
  # Odds ratios, 95% CI, and p-values for all covariates in the model
  conf.int <- Confint(mdl, exponentiate = TRUE)
  summary.mdl <- S(mdl,exponentiate = T)
  mdl.summary <- cbind(round(conf.int[2:nrow(conf.int),1], digits = 3),
                           cbind(paste(round(conf.int[2:nrow(conf.int),2], digits = 3), "-", round(conf.int[2:nrow(conf.int),3], digits = 3)),
                                 signif(summary.mdl$coefficients[2:nrow(conf.int),4], digits =3)))
  colnames(mdl.summary) <- c("OR", "CI 95%", "P-value")

  # PFAS/OH-PCBs OR, 95%CI, and p-value
  summary.PFAS[i,] <- mdl.summary[1,]
  
  # Covariate estimates and Anova p-values
  covariate.table[i,c(1,3,5)] <- format(signif(summary(mdl)$coefficients[3:5], digits = 3), scientific = T)
  covariate.table[i,c(2,4,6:7)] <- format(signif(Anova(mdl)[2:5,3], digits = 3), scientific = T)
}

write.csv(summary.PFAS, file = paste0(params$resultDir, "LogisticRegression/BaseModel/Males/summary.compounds.csv"))
write.csv(covariate.table, file = paste0(params$resultDir, "LogisticRegression/BaseModel/Males/covariates.csv"))
```

```{r  Females, message=FALSE}
covariate.table <- matrix(NA, nrow = 20, ncol = 7)
rownames(covariate.table) <- colnames(dt.pfas)[1:20]
colnames(covariate.table) <- c("Age - estimate", "Age - P",
                               "Current BMI - estimate", "Current BMI - P",
                               "BMI at 20 - estimate", "BMI at 20 - P",
                               "Week - P")

summary.PFAS <- matrix(NA, nrow = 20, ncol = 3)
colnames(summary.PFAS) <- c("OR", "CI 95%", "P-value")
rownames(summary.PFAS) <- colnames(dt.pfas)[1:20]

for(i in 1:20){
  temp <- dt.pfas %>% dplyr::select(append(colnames(dt.pfas)[i], c("Age", "Sex", "BMI_curr", "BMI_20", "Week", "status")))
  
  temp <- temp[which(temp$Sex == 1),] # Select only females
  names(temp)[1] <- "envCont"
  mdl <- glm(status ~ envCont + Age + BMI_curr + BMI_20 + Week, 
             data = temp, 
             family = binomial)
  
  # Odds ratios, 95% CI, and p-values for all covariates in the model
  conf.int <- Confint(mdl, exponentiate = TRUE)
  summary.mdl <- S(mdl,exponentiate = T)
  mdl.summary <- cbind(round(conf.int[2:nrow(conf.int),1], digits = 3),
                           cbind(paste(round(conf.int[2:nrow(conf.int),2], digits = 3), "-", round(conf.int[2:nrow(conf.int),3], digits = 3)),
                                 signif(summary.mdl$coefficients[2:nrow(conf.int),4], digits =3)))
  colnames(mdl.summary) <- c("OR", "CI 95%", "P-value")

  # PFAS/OH-PCBs OR, 95%CI, and p-value
  summary.PFAS[i,] <- mdl.summary[1,]
  
  # Covariate estimates and Anova p-values
  covariate.table[i,c(1,3,5)] <- format(signif(summary(mdl)$coefficients[3:5], digits = 3), scientific = T)
  covariate.table[i,c(2,4,6:7)] <- format(signif(Anova(mdl)[2:5,3], digits = 3), scientific = T)
}

write.csv(summary.PFAS, file = paste0(params$resultDir, "LogisticRegression/BaseModel/Females/summary.compounds.csv"))
write.csv(covariate.table, file = paste0(params$resultDir, "LogisticRegression/BaseModel/Females/covariates.csv"))
```

## Final model
```{r Males and Females, message=FALSE}
covariate.table <- matrix(NA, nrow = 20, ncol = 22)
rownames(covariate.table) <- colnames(dt.pfas)[1:20]
colnames(covariate.table) <- c("Age - estimate", "Age - P",
                               "Sex - estimate", "Sex - P",
                               "Current BMI - estimate", "Current BMI - P",
                               "BMI at 20 - estimate", "BMI at 20 - P",
                               "Regular smoking - estimate", "Regular smoking - P",
                               "Irregular smoking - estimate", "Irregular smoking - P",
                               "Number of children - estimate", "Number of children - P",
                               "History of infectious mononucleosis - estimate", "History of infectious mononucleosis - P",
                               "Sunhabit (high) - estimate", "Sunhabit (medium) - estimate", "Sunhabit - P",
                               "Swedish born - estimate", "Swedish born - P",
                               "Week - P")

summary.PFAS <- matrix(NA, nrow = 20, ncol = 3)
colnames(summary.PFAS) <- c("OR", "CI 95%", "P-value")
rownames(summary.PFAS) <- colnames(dt.pfas)[1:20]

for(i in 1:20){
  temp <- dt.pfas %>% dplyr::select(append(colnames(dt.pfas)[i], c("Age", "Sex", "BMI_curr", "BMI_20", "Week", "status",
                                                                   "reg.smoking", "irreg.smoking", "nr.children",
                                                                   "mono", "County", "sunhabit", "Swedish")))
  names(temp)[1] <- "envCont"
  mdl <- glmer(status ~ envCont + Age + Sex + BMI_curr + BMI_20 + reg.smoking + irreg.smoking + nr.children + mono + sunhabit  + Swedish + Week + (1|County), 
             data = temp, 
             family = binomial)
  
  # Odds ratios, 95% CI, and p-values for all covariates
  conf.int <- Confint(mdl, exponentiate = TRUE)
  summary.mdl <- S(mdl, exponentiate = T)
  mdl.summary <- cbind(round(conf.int[2:nrow(conf.int),1], digits = 3),
                           cbind(paste(round(conf.int[2:nrow(conf.int),2], digits = 3), "-", round(conf.int[2:nrow(conf.int),3], digits = 3)),
                                 signif(summary.mdl$fixed.effects[2:nrow(conf.int),4], digits =3)))
  colnames(mdl.summary) <- c("OR", "CI 95%", "P-value")

  # PFAS/OH-PCBs odds ratios, 95% CI, and p-values
  summary.PFAS[i,] <- mdl.summary[1,]
  
  # Covariate estimates and Anova p-values
  covariate.table[i,c(1,3,5,7,9,11,13,15,17,18,20)] <- format(signif(summary(mdl)$coefficients[3:13], digits = 3), scientific = T)
  covariate.table[i,c(2,4,6,8,10,12,14,16,19,21,22)] <- format(signif(Anova(mdl)[2:12,3], digits = 3), scientific = T)
}

write.csv(summary.PFAS, file = paste0(params$resultDir, "LogisticRegression/FinalModel/summary.compounds.csv"))
write.csv(covariate.table, file = paste0(params$resultDir, "LogisticRegression/FinalModel/covariates.csv"))
```
### Divided by sex
```{r Males, message=FALSE}
covariate.table <- matrix(NA, nrow = 20, ncol = 18)
rownames(covariate.table) <- colnames(dt.pfas)[1:20]
colnames(covariate.table) <- c("Age - estimate", "Age - P",
                               "Current BMI - estimate", "Current BMI - P",
                               "BMI at 20 - estimate", "BMI at 20 - P",
                               "Regular smoking - estimate", "Regular smoking - P",
                               "Irregular smoking - estimate", "Irregular smoking - P",
                               "History of infectious mononucleosis - estimate", "History of infectious mononucleosis - P",
                               "Sunhabit (high) - estimate", "Sunhabit (medium) - estimate", "Sunhabit - P",
                               "Swedish born - estimate", "Swedish born - P",
                               "Week - P")

summary.PFAS <- matrix(NA, nrow = 20, ncol = 3)
colnames(summary.PFAS) <- c("OR", "CI 95%", "P-value")
rownames(summary.PFAS) <- colnames(dt.pfas)[1:20]

for(i in 1:20){
  temp <- dt.pfas %>% dplyr::select(append(colnames(dt.pfas)[i], c("Age","BMI_curr", "BMI_20", "Week", "status",
                                                                   "reg.smoking", "irreg.smoking", 
                                                                   "mono", "County", "sunhabit", "Swedish")))
  temp <- temp[which(temp$Sex == 0),] # Select only males
  names(temp)[1] <- "envCont"
  mdl <- glmer(status ~ envCont + Age + BMI_curr + BMI_20 + reg.smoking + irreg.smoking + mono + sunhabit  + Swedish + Week + (1|County), 
             data = temp, 
             family = binomial)
  
  # Odds ratios, 95% CI, and p-values for all covariates
  conf.int <- Confint(mdl, exponentiate = TRUE)
  summary.mdl <- S(mdl, exponentiate = T)
  mdl.summary <- cbind(round(conf.int[2:nrow(conf.int),1], digits = 3),
                           cbind(paste(round(conf.int[2:nrow(conf.int),2], digits = 3), "-", round(conf.int[2:nrow(conf.int),3], digits = 3)),
                                 signif(summary.mdl$fixed.effects[2:nrow(conf.int),4], digits =3)))
  colnames(mdl.summary) <- c("OR", "CI 95%", "P-value")

  # PFAS/OH-PCBs odds ratios, 95% CI, and p-values
  summary.PFAS[i,] <- mdl.summary[1,]
  
  # Covariate estimates and Anova p-values
  covariate.table[i,c(1,3,5,7,9,11,13,14,16)] <- format(signif(summary(mdl)$coefficients[3:11], digits = 3), scientific = T)
  covariate.table[i,c(2,4,6,8,10,12,15,17,18)] <- format(signif(Anova(mdl)[2:10,3], digits = 3), scientific = T)
}

write.csv(summary.PFAS, file = paste0(params$resultDir, "LogisticRegression/FinalModel/Males/summary.compounds.csv"))
write.csv(covariate.table, file = paste0(params$resultDir, "LogisticRegression/FinalModel/Males/covariates.csv"))
```

```{r Females, message=FALSE}
covariate.table <- matrix(NA, nrow = 20, ncol = 20)
rownames(covariate.table) <- colnames(dt.pfas)[1:20]
colnames(covariate.table) <- c("Age - estimate", "Age - P",
                               "Current BMI - estimate", "Current BMI - P",
                               "BMI at 20 - estimate", "BMI at 20 - P",
                               "Regular smoking - estimate", "Regular smoking - P",
                               "Irregular smoking - estimate", "Irregular smoking - P",
                               "Number of children - estimate", "Number of children - P",
                               "History of infectious mononucleosis - estimate", "History of infectious mononucleosis - P",
                               "Sunhabit (high) - estimate", "Sunhabit (medium) - estimate", "Sunhabit - P",
                               "Swedish born - estimate", "Swedish born - P",
                               "Week - P")

summary.PFAS <- matrix(NA, nrow = 20, ncol = 3)
colnames(summary.PFAS) <- c("OR", "CI 95%", "P-value")
rownames(summary.PFAS) <- colnames(dt.pfas)[1:20]

for(i in 1:20){
  temp <- dt.pfas %>% dplyr::select(append(colnames(dt.pfas)[i], c("Age", "BMI_curr", "BMI_20", "Week", "status",
                                                                   "reg.smoking", "irreg.smoking", "nr.children",
                                                                   "mono", "County", "sunhabit", "Swedish")))
  temp <- temp[which(temp$Sex == 1),] # Select only females
  names(temp)[1] <- "envCont"
  mdl <- glmer(status ~ envCont + Age + BMI_curr + BMI_20 + reg.smoking + irreg.smoking + nr.children + mono + sunhabit  + Swedish + Week + (1|County), 
             data = temp, 
             family = binomial)
  
  # Odds ratios, 95% CI, and p-values for all covariates
  conf.int <- Confint(mdl, exponentiate = TRUE)
  summary.mdl <- S(mdl, exponentiate = T)
  mdl.summary <- cbind(round(conf.int[2:nrow(conf.int),1], digits = 3),
                           cbind(paste(round(conf.int[2:nrow(conf.int),2], digits = 3), "-", round(conf.int[2:nrow(conf.int),3], digits = 3)),
                                 signif(summary.mdl$fixed.effects[2:nrow(conf.int),4], digits =3)))
  colnames(mdl.summary) <- c("OR", "CI 95%", "P-value")

  # PFAS/OH-PCBs odds ratios, 95% CI, and p-values
  summary.PFAS[i,] <- mdl.summary[1,]
  
  # Covariate estimates and Anova p-values
  covariate.table[i,c(1,3,5,7,9,11,13,15,16,18)] <- format(signif(summary(mdl)$coefficients[3:12], digits = 3), scientific = T)
  covariate.table[i,c(2,4,6,8,10,12,14,17,19,20)] <- format(signif(Anova(mdl)[2:11,3], digits = 3), scientific = T)
}

write.csv(summary.PFAS, file = paste0(params$resultDir, "LogisticRegression/FinalModel/Females/summary.compounds.csv"))
write.csv(covariate.table, file = paste0(params$resultDir, "LogisticRegression/FinalModel/Females/covariates.csv"))
```